---
title: "Description and usage of MsBackendMsp"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of MsBackendMsp}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{MsBackendMsp}
    %\VignetteDepends{Spectra,BiocStyle,BiocParallel}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendMsp")`<br />
**Authors**: `r packageDescription("MsBackendMsp")[["Author"]] `<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(BiocStyle)
```

# Introduction

The `r Biocpkg("Spectra")` package provides a central infrastructure for the
handling of Mass Spectrometry (MS) data. The package supports interchangeable
use of different *backends* to import MS data from a variety of sources (such as
mzML files). The `MsBackendMsp` package adds support for files in NIST MSP
format which are frequently used to share spectra libraries. This vignette
illustrates the usage of the `MsBackendMsp` package and how it can be used to
import and export data in MSP file format.


# Installation

To install this package, start `R` and enter:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("MsBackendMsp")
```

This will install this package and all eventually missing dependencies.


# Importing MS/MS data from MSP files

The MSP file format allows to store MS/MS spectra (m/z and intensity of mass
peaks) along with additional annotations for each spectrum. A single MSP file
can thus contain a single or multiple spectra. Below we load the package and
define the file name of an MSP file which is distributed with this package.

```{r, message = FALSE}
library(MsBackendMsp)

nist <- system.file("extdata", "spectrum2.msp", package = "MsBackendMsp")
```

We next import the data into a `Spectra` object by specifying in the constructor
function the *backend* class which can be used to read the data (in our case a
`MsBackendMsp`).

```{r}
sp <- Spectra(nist, source = MsBackendMsp())
```

With that we have now full access to all imported spectra variables that can be
listed with the `spectraVariables` function.

```{r spectravars}
spectraVariables(sp)
```

Besides default spectra variables, such as `msLevel`, `rtime`, `precursorMz`, we
also have additional spectra variables such as the `name` or `adduct` that are
additional data fields from the MSP file.

```{r}
sp$msLevel
sp$name
sp$adduct
```

The NIST file format is however only loosely defined and variety of *flavors*
(or *dialects*) exist which define their own data fields or use different names
for the fields. The `MsBackendMsp` supports data import/export from all MSP
format variations by defining and providing different mappings between MSP data
fields and spectra variables. Also user-defined mappings can be used, which
makes import from any MSP flavor possible. Pre-defined mappings between MSP data
fields and spectra variables (i.e. variables within the `Spectra` object) are
returned by the `spectraVariableMapping` function.

```{r}
spectraVariableMapping(MsBackendMsp())
```

The names of this `character` vector represent the spectra variable names and
the values of the vector the MSP data fields. Note that by default, also all
data fields for which no mapping is provided are imported (with the field name
being used as spectra variable name).

This default mapping works well for MSP files from NIST or from other tools such
as MS-DIAL. MassBank of North America [MoNA](https://mona.fiehnlab.ucdavis.edu/)
however, uses a slightly different format. Below we read the first 6 lines of a
MSP file from MoNA.

```{r}
mona <- system.file("extdata", "minimona.msp", package = "MsBackendMsp")
head(readLines(mona))
```

The first 6 lines from a NIST MSP file:

```{r}
head(readLines(nist))
```

MSP files with MoNA flavor use slightly different field names, that are also not
all upper case, and also additional fields are defined. While it is possible to
import MoNA flavored MSP files using the default variable mapping that was used
above, most of the spectra variables would however not mapped correctly to the
respective spectra variable in the resulting `Spectra` object (e.g. the
precursor m/z would not be available with the expected spectra variable
`$precursorMz`).

The `spectraVariableMapping` provides however also the mapping for MSP files
with MoNA flavor.

```{r}
spectraVariableMapping(MsBackendMsp(), "mona")
```

Using this mapping in the data import will ensure that the fields get correctly
mapped.

```{r}
sp_mona <- Spectra(mona, source = MsBackendMsp(),
                   mapping = spectraVariableMapping(MsBackendMsp(), "mona"))
sp_mona$precursorMz
```

Note that in addition to the predefined variable mappings, it is also possible
to provide any user-defined variable mapping with the `mapping` parameter thus
enabling to import from MSP files with a highly customized format.

Multiple values for a certain spectrum are represented as duplicated fields in
an MSP file. The `MsBackendMsp` supports also import of such data. MoNA MSP
files use for example multiple `"Synon"` fields to list all synonyms of a
compound. Below we extract such values for two spectra within our `Spectra`
object from MoNA.

```{r}
sp_mona[29:30]$synonym
```

In addition to importing data from MSP files, `MsBackendMsp` allows also to
**export** `Spectra` to files in MSP format. Below we export for example the
`Spectra` with data from MoNA to a temporary file, using the default NIST MSP
format.

```{r}
tmpf <- tempfile()

export(sp_mona, backend = MsBackendMsp(), file = tmpf,
       mapping = spectraVariableMapping(MsBackendMsp()))
head(readLines(tmpf))
```

Or export the `Spectra` with data in NIST MSP format to a MSP file with MoNA
flavor.

```{r}
tmpf <- tempfile()

export(sp, backend = MsBackendMsp(), file = tmpf,
       mapping = spectraVariableMapping(MsBackendMsp(), "mona"))
head(readLines(tmpf))
```

Thus, this could also be used to convert between MSP files with different
flavors.


# Session information

```{r}
sessionInfo()
```
